---
title: "External Control Arm for an Early-Phase Oncology Trial"
subtitle: "A reproducible RWE walkthrough using propensity score methods"
author: "RWE Demo"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: show
    theme: cosmo
    embed-resources: true
bibliography: references.bib
csl: https://www.zotero.org/styles/vancouver
---

```{r setup}
#| include: false
library(ecasim)
library(dplyr)
library(ggplot2)
library(survival)
library(survminer)
library(gtsummary)
library(broom)

set.seed(2026)
theme_set(theme_minimal(base_size = 12))
```

# Background and objective

In early clinical development (Phase I / II), a single-arm trial design
is often used to accelerate decision-making in oncology. However,
without a randomized comparator it is difficult to attribute observed
outcomes to the investigational treatment rather than to patient
selection or natural disease history. **External control arms (ECA)**
constructed from real-world data (RWD) — such as electronic health
records (EHR), claims, or registries — provide a way to contextualize
single-arm trial results [@rahman2021fdaeca; @ema2020rwe].

This report walks through, end to end, how to build and analyze an ECA
using **propensity score matching** and **inverse probability of
treatment weighting (IPTW)**, with all code and data fully
reproducible.

::: {.callout-note}
## Learning objective
By the end of this report you should be able to:

1.  Recognize the sources of confounding when comparing a trial cohort
    to RWD.
2.  Estimate a propensity score, match patients, and compute IPT
    weights.
3.  Diagnose covariate balance before and after adjustment.
4.  Estimate and interpret a treatment effect (HR) on a time-to-event
    endpoint using Cox proportional hazards.
5.  Produce a regulator-style report that is fully reproducible.
:::

# Data: a synthetic but realistic RWD cohort

For teaching purposes we generate a synthetic dataset whose structure
mirrors an oncology RWD source. The trial cohort is drawn from the
**same** underlying population as the external controls but is
preferentially enriched for younger, fitter patients (lower ECOG, more
favorable biomarker, fewer prior lines). This is the kind of
selection bias one always faces when comparing trials to RWD.

```{r data}
cohort <- simulate_rwd(n_trial = 80, n_external = 800, seed = 42)
glimpse(cohort)
```

# Baseline characteristics

```{r tbl1}
cohort |>
  mutate(arm = factor(treatment, levels = c(0, 1),
                      labels = c("RWD control", "Trial"))) |>
  select(arm, age, sex, ecog, biomarker, prior_lines) |>
  tbl_summary(
    by = arm,
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    )
  ) |>
  add_p() |>
  add_overall() |>
  modify_caption("**Table 1. Baseline characteristics, unadjusted.**")
```

The trial arm is systematically different from the RWD pool. A naive
comparison of survival between the two arms would attribute part of
this baseline difference to the treatment. We need to adjust for
confounders.

# Propensity score estimation

A **propensity score** is the probability of being in the trial arm
given baseline covariates [@rosenbaum1983propensity]. Conditioning on
the propensity score balances those covariates between arms.

```{r ps}
weighted <- compute_iptw(cohort)
ggplot(weighted, aes(x = ps, fill = factor(treatment))) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("0" = "#4C78A8", "1" = "#F58518"),
                    labels = c("RWD control", "Trial"),
                    name = NULL) +
  labs(title = "Propensity score distribution by arm",
       x = "Estimated propensity score", y = "Density") +
  theme(legend.position = "top")
```

The two distributions overlap, which is the **positivity** assumption
needed for valid causal inference. Without overlap we cannot construct
a credible ECA in the regions of non-overlap.

# Approach 1: Propensity score matching (3:1)

We match each trial patient to up to 3 RWD controls within a caliper
of 0.2 SD on the logit-PS scale.

```{r match}
matched <- match_eca(cohort, ratio = 3, caliper = 0.2)
table(matched$treatment)
```

## Balance diagnostics

```{r balance}
balance_tbl <- function(d) {
  d |>
    mutate(arm = factor(treatment, levels = c(0,1),
                        labels = c("RWD control", "Trial"))) |>
    select(arm, age, sex, ecog, biomarker, prior_lines) |>
    tbl_summary(by = arm,
                statistic = list(all_continuous() ~ "{mean} ({sd})"))
}
balance_tbl(matched) |>
  modify_caption("**Table 2. Baseline characteristics after matching.**")
```

After matching, mean age, ECOG distribution and biomarker level are
much closer between arms.

# Approach 2: IPTW (stabilized weights)

```{r iptw-summary}
summary(weighted$siptw)
```

Extreme weights inflate variance. Truncation at the 1st/99th percentile
is a common stabilization.

```{r iptw-truncate}
q <- quantile(weighted$siptw, c(0.01, 0.99))
weighted$siptw_trunc <- pmin(pmax(weighted$siptw, q[1]), q[2])
```

# Treatment effect on overall survival

## Kaplan–Meier on the matched cohort

```{r km}
fit_km <- survfit(Surv(time_months, event) ~ treatment, data = matched)
ggsurvplot(fit_km, data = matched, risk.table = TRUE,
           conf.int = TRUE, palette = c("#4C78A8", "#F58518"),
           legend.labs = c("RWD control (matched)", "Trial"),
           xlab = "Months since index", ylab = "Overall survival")
```

## Cox proportional hazards — three estimators

```{r hr-table}
hr_naive   <- fit_cox(cohort)              |> mutate(method = "Unadjusted")
hr_matched <- fit_cox(matched)             |> mutate(method = "PS matching (3:1)")
hr_iptw    <- fit_cox(weighted, weights = weighted$siptw_trunc) |>
              mutate(method = "Stabilized IPTW")

bind_rows(hr_naive, hr_matched, hr_iptw) |>
  filter(term == "treatment") |>
  transmute(method,
            HR        = round(estimate, 2),
            `95% CI`  = sprintf("%.2f – %.2f", conf.low, conf.high),
            `p-value` = signif(p.value, 3)) |>
  knitr::kable(caption = "Estimated treatment effect on OS by method.")
```

The unadjusted HR is biased downward (artificially favorable to
treatment) because the trial arm starts healthier. After PS adjustment
the estimate is closer to the true generative HR of **0.70**.

# Sensitivity considerations

For a regulatory submission you would additionally:

- pre-specify the analysis plan and lock the protocol;
- assess unmeasured confounding via the **E-value** [@vanderweele2017evalue];
- run quantitative bias analyses;
- repeat with alternate matching ratios, calipers, and weight
  truncation thresholds;
- assess the **proportional hazards** assumption (Schoenfeld residuals);
- consider doubly-robust estimators (e.g., AIPW, TMLE).

# Reproducibility

```{r repro}
sessionInfo()
```

The exact R version and package versions are pinned via the
`.devcontainer` and the package `DESCRIPTION`.  Every figure and table
above is regenerated from the same seed (`42`) every time the report is
rendered — so two reviewers running the report on different machines
get the same numbers.

# References {.unnumbered}

::: {#refs}
:::
